---
title: "The easyclimate R package: easy access to high-resolution daily climate data for Europe"

# subtitle: "Short title: easyclimate R package"

author: "Verónica Cruz-Alonso (ORCID: XXXX) ^1,2^ §, 
    Christoph Pucher (ORCID: ZZZZ) ^3^ §, 
    etc etc,
    Francisco Rodríguez-Sánchez (ORCID: 0000-0002-7981-1599) ^10,11^"

# date: `r Sys.Date()`

output:
  bookdown::word_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_md: no
    number_sections: no
    toc: no
    reference_docx: Word_template.docx
    
bibliography: 
  - references.bib
  - knitcitations.bib
csl: journal-of-vegetation-science.csl
# csl: https://www.zotero.org/styles/ecology

---

^1^ UAH

^2^ Harvard

^3^ Boku

^10^ Departamento de Biología Vegetal y Ecología, Universidad de Sevilla, Spain.

^11^ Estación Biológica de Doñana, Consejo Superior de Investigaciones Científicas, Spain.

§ These authors share the first-author position.

Correspondence:

Verónica Cruz-Alonso, Harvard Univ, Boston, USA.
Email: veronica.cral@gmail.com


# FUNDING INFORMATION

FRS was supported by the VI Plan Propio de Investigación of Universidad de Sevilla (VI PPIT-US).


# ABSTRACT

Write your abstract here (200 WORDS MAXIMUM)


*Keywords*: R package, climate, Europe, reproducibility, cloud-optimised geotiffs, (8-12 KEYWORDS)



```{r setup, include=FALSE, cache=FALSE, message = FALSE}

library("knitr")

### Chunk options: see http://yihui.name/knitr/options/ ###
## Text results
opts_chunk$set(echo = FALSE, warning = TRUE, message = TRUE)

## Code decoration
opts_chunk$set(tidy = TRUE, comment = NA, highlight = TRUE)

## Cache
# opts_chunk$set(cache = 2, cache.path = "output/cache/")

## Plots
opts_chunk$set(fig.path = "output/figures/", dpi = 300)

```


```{r knitcitations, cache = FALSE}
library(knitcitations)
cleanbib()   
cite_options(citation_format = "pandoc")
```




# INTRODUCTION

Write your introduction here. You can cite bibliography like this [@Yan2011; @Sutherland2011] if you provide a `BibTeX` file with references. See http://rmarkdown.rstudio.com for more information. 

You can even specify the desired output format for your bibliography by including a style file for a specific journal (e.g. "ecology.csl"). Many different bibliography styles (CSL files) can be obtained at http://citationstyles.org/, https://github.com/citation-style-language/styles, or https://zotero.org/styles.



# METHODS


## Study Area

We worked in a **beautiful** place with lots of trees, like *Quercus suber* and *Laurus nobilis*.


## Data collection and analysis

```{r read_data, eval = FALSE}
dataset <- read.csv("mydata.csv")
```

We applied a linear model where

$$
y_{i} = \alpha + \beta*x_{i} 
$$

```{r model, eval = FALSE}
model <- lm(y ~ x)
```

We used the statistical language `R` `r citep(citation())` for all our analyses. These were implemented in dynamic rmarkdown documents using `knitr` `r citep(citation("knitr"))` and `rmarkdown` `r citep(citation("rmarkdown"))` packages. All the multilevel models were fitted with `lme4` `r citep(citation("lme4"))`. 




# RESULTS

Trees in forest *A* grew taller than those in forest *B* (mean height: `r mean(25, 31, 28)` versus `r mean(13, 19, 16)` m). 

And many more cool results that get updated dynamically, e.g. see Table \@ref(tab:Table-mtcars) and Fig. \@ref(fig:Fig-landscape). Note Tables and Figures are cross-linked and numbered automatically. 


# DISCUSSION

Discuss.



# CONCLUSIONS

Wrap up


# ACKNOWLEDGEMENTS

On the shoulders of giants.


# AUTHOR CONTRIBUTIONS

Use CREDIT taxonomy of author contributions: https://rollercoaster.shinyapps.io/tenzing/

# DATA AVAILABILITY STATEMENT

The climate datasets are publicly available at ftp://palantir.boku.ac.at/Public/ClimateData/. The easyclimate R package is available on CRAN (url) and GitHub (https://github.com/VeruGHub/easyclimate).


# REFERENCES

```{r write_citations, cache=FALSE, include=FALSE}
write.bibtex(file = "knitcitations.bib")
```

<div id ="refs"></div>


\newpage

# Table (on new page)

```{r Table-mtcars}
kable(mtcars[10:16, ], caption = "Now a subset of mtcars dataset.")
```


\newpage

# Figure (on new page)

```{r Fig-landscape, fig.cap="A boxplot."}
a <- sort(rnorm(100))
b <- c(rep("Group Small", 35), rep("Group Big", 65))
boxplot(a ~ b)
```



